Finiteness in Cubical Type Theory

### 2020-06-25 sections/introduction.tex

\chapter{Introduction}
We are interested in constructive notions of finiteness, formalised in Cubical
Type Theory~\cite{cohenCubicalTypeTheory2016}.
In this paper we will explore five such notions of finiteness, including their
categorical interpretation, and use them to build a simple proof-search library
facilitated in a fundamental way by univalence.
Along the way we will use the Countdown
problem~\cite{huttonCountdownProblem2002} as an example, and provide a program
which produces verified solutions to the puzzle.
We will also briefly examine countability, and demonstrate its parallels and
differences with finiteness.
\section{The Varieties of Finiteness}
In Section~\ref{finiteness-predicates} we will explore a number of different
predicates for finiteness.
In contrast to classical finiteness, in a constructive setting there are many
predicates which all have some claim to being the formal interpretation of
``finiteness''~\cite{coquandConstructivelyFinite2010}.
The particular predicates we are interested in are organised in
Figure~\ref{finite-classification}: each arrow in the diagram represents a proof
that one predicate can be derived from another.
Each arrow in Figure~\ref{finite-classification} corresponds to a proof of
implication: cardinal finiteness, for instance, with a strict total order,
implies split enumerability (Theorem~\ref{cardinal-to-manifest-bishop}).

@@

Theoretically speaking, showing that sets in Homotopy Type Theory form a topos
(with some caveats) is an important step in characterising the categorical
implications of Homotopy Type Theory, first proven
in~\cite{rijkeSetsHomotopyType2015}.
Our work is a formalisation of this result (and the first such formalisation
that we are aware of).
The proof that decidable Kuratowski finite sets form a \(\Pi\)-pretopos is
additional to that.
\section{Countability Predicates}
After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates
(Section~\ref{infinite}).
We will see that we lose closure under function arrows, but we gain it under the
Kleene star (Theorem~\ref{split-countability-sigma}).
\section{Search}
All of our work is formalised in Cubical
Agda~\cite{vezzosiCubicalAgdaDependently2019}: as a result, the constructive
interpretation of each proof is actually a program which can be run on a
computer.
In finiteness in particular, these programs are particularly useful for
exhaustive search.

>>>>

\chapter{Introduction}
\todo{regroup into old structure}
We are interested in constructive notions of finiteness, formalised in Cubical
Type Theory~\cite{cohenCubicalTypeTheory2016}.
In this paper we will explore five such notions of finiteness, including their
categorical interpretation, and use them to build a simple proof-search library
facilitated in a fundamental way by univalence.
Along the way we will use the Countdown
problem~\cite{huttonCountdownProblem2002} as an example, and provide a program
which produces verified solutions to the puzzle.
We will also briefly examine countability, and demonstrate its parallels and
differences with finiteness.
\section{The Varieties of Finiteness}
\todo{Make all references parenthetical}
In Section~\ref{finiteness-predicates} we will explore a number of different
predicates for finiteness.
In contrast to classical finiteness, in a constructive setting there are many
predicates which all have some claim to being the formal interpretation of
``finiteness''~\cite{coquandConstructivelyFinite2010}.
The particular predicates we are interested in are organised in
Figure~\ref{finite-classification}: each arrow in the diagram represents a proof
that one predicate can be derived from another.
Each arrow in Figure~\ref{finite-classification} corresponds to a proof of
implication: cardinal finiteness, for instance, with a strict total order,
implies split enumerability (Theorem~\ref{cardinal-to-manifest-bishop}).

@@

Theoretically speaking, showing that sets in Homotopy Type Theory form a topos
(with some caveats) is an important step in characterising the categorical
implications of Homotopy Type Theory, first proven
in~\cite{rijkeSetsHomotopyType2015}. \todo{This reference should be citet not citep}
Our work is a formalisation of this result (and the first such formalisation
that we are aware of).
The proof that decidable Kuratowski finite sets form a \(\Pi\)-pretopos is
additional to that.
\section{Countability Predicates}
After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates
(Section~\ref{infinite}).
We will see that we lose closure under function arrows, but we gain it under the
Kleene star (Theorem~\ref{split-countability-sigma}).
\section{Search}
All of our work is formalised in Cubical
Agda~\cite{vezzosiCubicalAgdaDependently2019}: as a result, the constructive
interpretation of each proof is actually a program which can be run on a
computer.
In finiteness in particular, these programs are particularly useful for
exhaustive search.
